---
title: "Gridded Analysis"
---

```{r, include = FALSE}
suppressPackageStartupMessages(library("eupp"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("stars"))
suppressPackageStartupMessages(library("rjson"))
dir.create("_cache", showWarnings = FALSE)
library("rjson")
```

This example shows how to download gridded analysis data. The data set
is based on ECMWFs latest ERA5 reanalysis. For details on the basic use
of _eupp_ gridded functionality please have a look at the
[Getting started](gridded.html) article.

Gridded analysis data (based on ECMWFs ERA5 reanalysis) is available
for a series of surface and pressure level fields introcuded below.


## Surface data


### Parameters

```{r gridded_analysis_surface_table, echo = FALSE}
conf <- eupp_config("analysis", "surface", date = "2017-01-01", step = 5, cache = "_cache")
eupp_download_gridded(conf, "_test.grb", "grib", overwrite = TRUE)
tmp <- system("grib_ls -j -p shortName,name _test.grb", intern = TRUE)
tmp <- fromJSON(paste(tmp, collapse = ""))
tmp <- do.call(rbind, lapply(tmp[[1]], as.data.frame))
tmp <- data.frame(product = conf$product, level = conf$level,
                  param = tmp$shortName, description = tmp$name)
knitr::kable(tmp)
```

### Available steps

When downloading analysis date, the argument `steps` (`eupp_config()`)
defines the time of the day of the data to be processed.

```{r gridded_analysis_surface_request}
conf <- eupp_config("analysis", "surface",
                    param = c("2t", "cp"),
                    date  = "2018-07-01",
                    steps = 0:23,          # 00 to 23 UTC
                    cache = "_cache")

# Loading inventory for a quick check
inv <- eupp_get_inventory(conf)
table(inv$valid)
```

### Downloading GRIB data

For details how to download the data defined by the configuration above
please visit the [Getting started](gridded.html) article. As there
are some specialities when handling analysis data here is a short example.

Therefore we are downloading analysis data for three consecutive days (could be
done for e.g., a month). Again surface data, 2m temperature and total precipitation.

```{r gridded_analysis_surface_request2}
conf2 <- eupp_config("analysis", "surface",
                     param = c("2t", "cp"),
                     date  = as.Date("2018-07-01") + 0:1,
                     steps = 0:23,          # 00 to 23 UTC
                     cache = "_cache")

# Downloading grib file
eupp_download_gridded(conf2, "_analysis.grb", overwrite = TRUE)
```

The _eupp_ package downloads the required GRIB messages even if they are spread
over multiple files on the data store - and puts them into one local file.

### Interpolating data

We are interested in bilinearely interpolated values for the following two
locations:

```{r}
library("sf")
locations <- data.frame(name = c("Innsbruck", "Brussels"),
                        lon  = c(11.39, 4.35),
                        lat  = c(47.27, 50.85))
(locations <- st_as_sf(locations, coords = c("lon", "lat"), crs = 4326))
```

Default use:

```{r analysis_interpolate_one}
ip1 <- eupp_interpolate_grib("_analysis.grb", at = locations,
                             atname = "name", bilinear = TRUE)
head(ip1, n = 4)
```

**Note:** When looking at the first two lines one can see that the same
`valid` time stamp (when the analysis is valid) occurs twice. The reason
is that the 2m temperature is a 'direct' field from the analysis process,
total precipitation (as an example) is a short-term forecast and thus
initialized earlier and showing a forecast step.

We can get around this by setting `ignore_init = TRUE`. _eupp_ now combines
the data on the `valid` time stamp only. Spot the difference:

```{r analysis_interpolate_two}
ip2 <- eupp_interpolate_grib("_analysis.grb", at = locations,
                             atname = "name", bilinear = TRUE,
                             ignore_init = TRUE)
head(ip2, n = 4)
```

We will, however, lose some meta information (compared to the version above).
In case this is important `wide = FALSE` returns one line in the data.frame
for each message in the GRIB file alongside with all the loaded meta information 
(in this case `ignore_init` has no effect).

```{r analysis_interpolate_three}
ip3 <- eupp_interpolate_grib("_analysis.grb", at = locations,
                             atname = "name", bilinear = TRUE,
                             wide = FALSE)
head(ip3, n = 4)
```

This can be handy in some situations.

```{r gridded_analysis_lineplot, fig = TRUE, fig.width = 10, fig.height = 3.5, out.width = "100%"}
library("ggplot2")
ggplot(ip3) +
       geom_line(aes(x = valid, y = value, color = name)) +
       facet_wrap("param", scales = "free_y")
```




## Pressure level data

### Parameters

```{r gridded_analysis_pressure_table, echo = FALSE}
conf <- eupp_config("analysis", "pressure", date = "2017-01-01", step = 5, cache = "_cache")
eupp_download_gridded(conf, "_test.grb", "grib", overwrite = TRUE)
tmp <- system("grib_ls -j -p level,shortName,name _test.grb", intern = TRUE)
tmp <- fromJSON(paste(tmp, collapse = ""))
tmp <- do.call(rbind, lapply(tmp[[1]], as.data.frame))
tmp <- data.frame(product = conf$product, level = conf$level,
                  param = sprintf("%s (%d)", tmp$shortName, tmp$level),
                  description = tmp$name)
knitr::kable(tmp)
```

### Available steps

When downloading analysis date, the argument `steps` (`eupp_config()`)
defines the time of the day of the data to be processed.

```{r gridded_analysis_pressure_request}
conf <- eupp_config("analysis", "pressure",
                    date  = "2018-07-01",
                    steps = 0:23,
                    cache = "_cache")
inv <- eupp_get_inventory(conf)
table(inv$valid)
table(inv$levelist, inv$param)
```


### Downloading GRIB data

Same as for the 'surface analysis' variable shown in the previous sections.
For demonstration the configuration is adjusted to download pressure analysis
for two consecutive days (all parameters).

```{r gridded_analysis_pressure_request2}
conf2 <- eupp_config("analysis", "pressure",
                     date  = as.Date("2018-07-01") + 0:1,
                     steps = 0:23,          # 00 to 23 UTC
                     cache = "_cache")

# Downloading grib file
eupp_download_gridded(conf2, "_pressure.grb", overwrite = TRUE)
```

### Interpolating data

Interpolating the data for the the same two locations (see object `locations` above).
Basic use:


```{r analysis_interpolate_four}
ip4 <- eupp_interpolate_grib("_pressure.grb", at = locations,
                             atname = "name", bilinear = TRUE)
head(ip4, n = 4)
```

As the current pressure level data set does not contain any short-term forecasts
the wide format is not sparse. In case needed, the same functionality is available.
E.g., getting the long format for easy plotting ...


```{r analysis_interpolate_five}
ip5 <- eupp_interpolate_grib("_pressure.grb", at = locations,
                             atname = "name", bilinear = TRUE,
                             wide = FALSE)
head(ip5, n = 4)
```

```{r gridded_pressure_lineplot, fig = TRUE, fig.width = 10, fig.height = 6, out.width = "100%"}
ggplot(ip5) +
       geom_line(aes(x = valid, y = value, color = name)) +
       facet_wrap("param", scales = "free_y")
```


