---
title: "Getting started with gridded data"
---

```{r, include = FALSE}
suppressPackageStartupMessages(library("eupp"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("stars"))
```

This _eupp_ package provides access to a variety of gridded data sets provided
within the scope of the European Post-Processing benchmark project.

The gridded data sets consists of (re)analysis data used as the gridded ground-trough
in some scenarios, deterministic and ensemble forecasts for the training and test
period defined within the project as well as hindcasts (or reforecasts) to be worked
with. While the different data sets differ in form and extent, the _eupp_ package
provides a uniform interface to download and--to some extent--process the data.

## Purpose of this article

This article (_Getting started with gridde data_) shows the main use the
`eupp_*_gridded()` functionality with some minimal examples on how to working
with the data. Therefore, different types of gridded data sets will be used
in different situations. Dedicated articles are available highlighting 
specific characteristics and explicit examples for the different types of data.
Namely:

* [Gridded analysis data](gridded_analysis.Rmd)
* [Gridded forecast data](gridded_forecasts.Rmd)
* [Gridded hindcast data](gridded_hindcast.Rmd) (reforecasts)

Note that these articles will often refer back to this 'getting started' article
as most functions/procedures work the very same independent of the type of gridded
data.



## Underlying concept {#gridded-concept}

The data set has been designed and prepared by colleagues at the
[RMI](https://climdyn.meteo.be/) in Brussels (part of part of the R&D Department
of the Royal Meteorological Institute of Belgium).  The gridded data set
consists of different [ECMWF](https://www.ecmwf.int) products (see
[LICENSE](https://github.com/retostauffer/eupp/blob/main/DATA_LICENSE)) with
access granted via the `europeanweather.cloud` S3 bucket.

All gridded data sets are stored as GRIB version 1 files, alongside with
a GRIB index file. These files can technically be accessed directly, however,
this may be inconvenient for most/some. Thus, the _eupp_ package provides
an interface to download the data.


![Rough scheme on the download/processing process.](images/download_grib.svg)

Independent of the product or subset, the procedure for all products is the same:

1. Define what data should be downloaded (`eupp_config()`).
2. Download/retrieve the data (GRIB version 1, NetCDF, [`stars`][stars].

This article contains a series of links to the article
"[_Gridded data: Advanced_](gridded_advanced.html)" not required to follow as
casual users but might be helpful to show some insights to more advanced users,
programmers, and supporters.

<div style="color: gray;">
Under the hood, the _eupp_ package performs a series of intermediate steps
for (2) to achive the goal. 

a.  Defining the GRIB index files required to identify the necessary GRIB messages
b.  Downloading and parsing the GRIB index files to identify files and byte ranges
c.  Partially downloading the GRIB files (required messages via `curl`) and stores
    the requested messages in a new GRIB version 1 file.
d.  If a NetCDF file has been requested: making required manipulations on the GRIB
    file and converting it to NetCDF, wherefore [ecCodes][ecCodes] needs to be installed.
e.  If a [`stars`][stars] object has been requested:
    read the NetCDF file. This goes trough the intermediate step of creating a NetCDF
    file; thus [ecCodes][ecCodes] is necessary.
</div>


## Define dataset to be downloaded {#gridded-config}

Step one
Before starting downloading data, a configuration object must be created using
`eupp_config()` which contains the specification of the data to be retrieved.


```{r, include = FALSE}
dir.create("_cache", showWarnings = FALSE) # Not mandatory
```

```{r}
# Loading the package
library("eupp")

# Create custom configuration
conf <- eupp_config(product   = "forecast",
                    level     = "surf",
                    type      = "ens",
                    date      = "2017-07-01",
                    parameter = c("cp", "2t"),
                    steps     = c(24L, 240L), # +1 and +10 days ahead
                    cache     = "_cache")     # optional; caching grib index
```

## Getting inventory {#gridded-inventory}

Typically not done by the end-user but handy to see what messages will be downloaded
or to have a look at available messages before downloading the data itself is to look
at the GRIB inventory.

```{r}
inv <- eupp_get_inventory(conf)
head(inv)
dim(inv)
```

In this case the configuration (`conf`) defines a set of `r nrow(inv)` messages to
be processed/downloaded. To see what messages are available, one can simply set up
a configuration for a specific `product`/`level`/`type`/`date` but not specifying
`steps` or `parameters`. This will return the full inventory with all available
parameters and steps.



## Getting data: GRIB format {#gridded-download-grib}

From `eupp_get_inventory()` we know that there are `r nrow(inv)` fields matching
our configuration. `eupp_download_gridded()` allows us to retrieve the data in
the original GRIB version 1 file format by specifying `output_format = "grib"`.

The function will first download/parse the GRIB index file (uses `cache` if specified)
to know which GRIB messages are required given the configuration (`conf`) before starting
to download the requires messages. All messages matching the configuration will be stored
in one single file specified by `output_file` (GRIB version 1 file format).

```{r}
eupp_download_gridded(conf, output_file = "_test.grb", overwrite = TRUE)
```

Alongside with the GRIB vile (`"_test.grb"`) an `.rds` file `"_test.grb.rds"` will
be stored containing the GRIB inventory (meta information about the fields). Whilst
not really required this allows to interpolate the GRIB files without the need
to have [ecCodes][ecCodes] to be installed (see next section).

## Interpolate GRIB files

The _eupp_ package allows to interpolate GRIB data directly. Commonly this is done
using additional libraries which are able to read the GRIB meta information (index)
such as the [ecCodes][ecCodes].

[`stars`][stars] can also read GRIB files directly (via [`rgdal`][rgdal]), it does,
however, not return this meta information. `eupp_interpolate_gridded()` thus does
the following:

* Checks if the `.rds` file exists alongside the GRIB file to be interpolated (see previous section).
  Uses this information to perform interpolation (does not require [ecCodes][ecCodes]).
* If the `.rds` does not exist, `grib_ls` ([ecCodes][ecCodes]) is called to create
  the inventory/index from the GRIB file.

Currently, `eupp_interpolate_gridded()` only allows to interpolate one or multiple
points (`POINT` features). The interpolation is performed via [`stars`][stars] before
being manipulated and brought to a 'more usable' form.

##### Point locations

First an `sf` object containing the target locations has to be created. Only point
locations are allowed and the object must have a valid coordinate reference system (CRS).

```{r}
library("sf")
locations <- data.frame(name = c("Innsbruck", "Brussels"),
                        lon  = c(11.39, 4.35),
                        lat  = c(47.27, 50.85))
(locations <- st_as_sf(locations, coords = c("lon", "lat"), crs = 4326))
```

Once available, the GRIB file can be interpolated.

```{r}
ip <- eupp_interpolate_grib("_test.grb", at = locations,
                            atname = "name", bilinear = TRUE)
```

The warnings come from `readGDAL()` ([`rgdal`][rgdal]) and can be ignored
at this point. By default, a wide-format is returned, but a long format
can be retrieved if needed.

```{r}
head(ip[, 1:11]) # First 11 columns only

# Long format; contains more extensive information
# (differs between rds/grib_ls).
head(eupp_interpolate_grib("_test.grb", at = locations,
                           atname = "name", wide = FALSE), n = 3)
```

Please check out the additional arguments of `eupp_interpolate_grib()` for details
on the arguments and additional arguments not demonstrated here.


## Further functionality

The _eupp_ contains some additional functionality to download/process gridded
data sets. They, however, all go trough `grib_to_netcdf` ([ecCodes][ecCodes])
which comes with a series of benefits and drawbacks. A separate article
shows that, however, when using it keep in mind that this must be seen
as 'experimental'.

* See: [Additional functionality](gridded_additional_functionality.html)



```{r gridded_cleaning_up, include = FALSE}
if (dir.exists("_cache")) unlink("_cache")
if (file.exists("_test.nc")) unlink("_test.nc")
if (file.exists("_test.grb")) unlink("_test.grb")
if (file.exists("_test.grb.rds")) unlink("_test.grb.rds")
```


[ecCodes]: https://confluence.ecmwf.int/display/ECC/ecCodes+Home
[stars]: https://cran.r-project.org/package=stars
[netcdf4]: https://cran.r-project.org/package=ncdf4
[rgdal]: https://cran.r-project.org/package=rgdal
