---
title: "Getting gridded data"
---


![Rough scheme on the download/processing process.](images/download_grib.svg)

The data sets are stored as GRIB version 1 files on the servers alongside
with a GRIB index file which contains the inventory. The inventory contains
information of all fields inside the GRIB file and allows to partially
download the fields.

When downloading data, the package always first accesses the index file
via `eupp_get_inventory()` before starting to download the data.
`eupp_download_gridded()` will download (parts) of the GRIB file and
either store it as GRIB, or converts it to NetCDF (requires NetCDF support and ecCodes
to be installed; see installation notes).

`eupp_get_gridded()` interfaces `eupp_download_gridded()` but deletes the file
after use and returns an _R_ object. This can be handy for small tasks or checks,
or if you want to store the data as _R_ objects yourself. Also requires NetCDF support.


## Specify data to be downloaded

Before starting downloading data, a configuration object must be created using
`eupp_config()` which contains the specification of the data to be retrieved.

```{r, include = FALSE}
suppressPackageStartupMessages(library("eupp"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("stars"))
```

```{r, include = FALSE}
dir.create("_cache", showWarnings = FALSE) # Not mandatory
```

```{r}
# Loading the package
library("eupp")

# Create custom configuration
conf <- eupp_config(product   = "forecast",
                    level     = "surf",
                    type      = "ens",
                    date      = "2017-07-01",
                    parameter = c("cp", "2t"),
                    steps     = c(24L, 240L), # +1 and +10 days ahead
                    cache     = "_cache")     # optional; caching grib index
```

## Get full file URL

If you are interested where the data is stored, you can make use of the
function `eupp_get_source_urls()`. Typically not required to be called by an end-user.

```{r}
urls <- eupp_get_source_urls(conf)                           # GRIB file
urls <- eupp_get_source_urls(conf, fileext = "index")        # GRIB index/inventory
```

## Getting GRIB inventory

Given the configuration object `conf` from above, we can now retrieve the GRIB
inventory which matches our requirements.

```{r}
inv <- eupp_get_inventory(conf)
dim(inv)
head(inv, n = 3)
```

As we have set `cache` when calling `eupp_config()` the GRIB index file is
stored locally. If the same index file is required once again the loading times
should be much faster as the cached file is reused. Not required but may save
some time in certain situations.


## Getting data: GRIB format

From `eupp_get_inventory()` we know that there are `r nrow(inv)` fields matching
our configuration. `eupp_download_gridded()` allows us to retrieve the data in
the original GRIB version 1 file format by specifying `output_format = "grib"`.

The function will first download/parse the GRIB index file (uses `cache` if specified)
to know which GRIB messages are required given the configuration (`conf`) before starting
to download the requires messages. All messages matching the configuration will be stored
in one single file specified by `output_file` (GRIB version 1 file format).

```{r}
eupp_download_gridded(conf, output_file = "_test.grb", output_format = "grib", overwrite = TRUE)
```

## Getting data: NetCDF

Alternatively, the function allows to specify `output_format = "nc"` to store
the data in the NetCDF file format. The procedure is, in large parts, similar
to retrieving the data in GRIB format (see above). The function first downloads
and stores the data in a temporary GRIB file before calling
[`grib_to_netcdf`](https://confluence.ecmwf.int/display/ECC/grib_to_netcdf); a console
tool provided by ECMWFs [ecCodes](https://confluence.ecmwf.int/display/ECC/ecCodes+Home).

There seem to be ways to use `ecCodes` under Windows, see
[here](https://github.com/nawendt/gribr#windows-install-options) or
[here](https://www.ecmwf.int/en/newsletter/159/news/eccodes-and-magics-available-under-windows).

`grib_to_netcdf` is called with option `-k 3` resulting in a file with the
'netCDF-4 file format'.  Can be changed by adjusting the argument `netcdf_kind`
(default `3`, see `grib_to_netcdf` documentation).
Note that the conversion from GRIB to NetCDF adjusts the
variable names if needed (e.g., `2t` gets `t2m`).

```{r}
eupp_download_gridded(conf, output_file = "_test.nc", output_format = "nc", overwrite = TRUE)
```

These files can later be processed with the packages/software of your choice
(e.g., via [`ncdf4`](https://cran.r-project.org/package=ncdf4),
[`stars`](https://cran.r-project.org/package=stars), CDS, ...).
Some examples how to proceed from here ...

##### Reading data using `ncdf4`

```{r}
library("ncdf4")
nc <- nc_open("_test.nc")
nc
```

##### Reading data using `stars`

```{r demo_read_stars, eval = FALSE}
library("stars")
x <- read_stars("_test.nc")
```

```{r demo_read_stars_hidden, include = FALSE}
library("stars")
x <- read_stars("_test.nc")
```

```{r demoplot_stars, fig = TRUE, fig.width = 10, fig.height = 5, out.width = "100%"}
print(x)
dim(x)
names(x)
st_get_dimension_values(x, "time") # date/time when valid

# Plotting 2m dry air temperature
plot(x["t2m"],
     main   = "t2m",
     col    = hcl.colors(21, "Blue-Red 2", rev = TRUE),
     breaks = seq(min(x$t2m, na.rm = TRUE), max(x$t2m, na.rm = TRUE), length = 22))

# Plotting convective precipitation (mm)
bk <- unique(quantile(x[["cp"]], p = seq(0, 1, by = 0.05)))
plot(x["cp"] * 1e3,
     main   = "cp",
     col    = hcl.colors(length(bk) - 1, "Purple-Yellow", rev = TRUE),
     breaks = bk * 1e3)
```

An alternative way to visualize the data using [`ggplot2`](https://cran.r-project.org/package=ggplot2),
[`colorspace`](https://cran.r-project.org/package=colorspace) and 
[`sf`](https://cran.r-project.org/package=sf) as well as 
[`rnaturalearth`](https://cran.r-project.org/package=rnaturalearth) for outlines:
```


```{r demoplot_ggplot, fig = TRUE, fig.width = 10, fig.height = 7, out.width = "60%", fig.align = "center"}
library("sf")
library("ggplot2")
library("colorspace")
library("rnaturalearth")

ne <- ne_countries(continent = "europe", returnclass = "sf")
st_crs(x) <- st_crs(ne)

# Plotting convective precipitation; member number 10.
# x["cp",,,number_idx,] is subsetting the stars object.
number_idx <- which(st_get_dimension_values(x, "number") == 10)
ggplot() + geom_stars(data = x["cp",,,number_idx,] * 1e3) +
           facet_wrap("time") +
           scale_fill_continuous_sequential("Purple-Yellow") +
           geom_sf(data = ne, fill = "transparent") +
           coord_sf(xlim = c(-6, 17), ylim = c(36, 67), expand = FALSE)
```

```{r demoplot_ggplot_lambert, fig = TRUE, fig.width = 10, fig.height = 7, out.width = "80%", fig.align = "center"}
# CRS ESRI:102014: Europe_Lambert_Conformal_Conic projection
# Hardcoded x/y-limits added; limit to southern Europe.
tmp <- st_transform(x, crs = st_crs("ESRI:102014"))
ggplot() + geom_stars(data = tmp["t2m",,,1,2]) +
           scale_fill_continuous_sequential("Red-Yellow") +
           geom_sf(data = ne, fill = "transparent") +
           coord_sf(xlim = c(-1.2e6, 6e5), ylim = c(7e5, 2e6), expand = TRUE)
```


```{r gridded_cleaning_up, include = FALSE}
if (dir.exists("_cache")) unlink("_cache")
if (file.exists("_test.nc")) unlink("_test.nc")
if (file.exists("_test.grb")) unlink("_test.grb")
```




