---
title: "Additional/experimental functionality"
---

```{r, include = FALSE}
suppressPackageStartupMessages(library("eupp"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("stars"))
```

This article shows additional _eupp_ functionality for
handling gridded data. This must seen as 'experimental'
as it not always delivers the data as one might expect.

This is mainly related to the fact that everything included
in this article is based on files in the NetCDF format using
`grib_to_netcdf` ([ecCodes][ecCodes]) to convert the GRIB version 1
files to NetCDF and then processed the data using [`stars`][stars].

The problem is that the NetCDF standard does not support all the
required dimensions (e.g., initial time + step; levels; ensemble members).
which can cause issues when:

* Initialization dates and steps overlap (non-unique times)
* Dimensions are no longer required (e.g., downloading ensemble member 10
    only: the demension, and thus the information, will be lost).

Thus, this should not be used in production. Functionality may change
in the future if someone has a great idea about a general workaround
for all these issues.

Else it is suggested to only download grib data and then use custom
software/the software of your choice to properly process everythign.


## Setting up

In this article functionality is shown which require an _eupp_
configuration object and an `sf` object with points of interest.
Thus, these are set up first:

```{r, include = FALSE}
dir.create("_cache", showWarnings = FALSE) # Not mandatory
```

```{r}
library("eupp")

# Create custom configuration
conf <- eupp_config(product   = "forecast",
                    level     = "surf",
                    type      = "ens",
                    date      = "2017-07-01",
                    parameter = c("cp", "2t"),
                    steps     = c(24L, 240L), # +1 and +10 days ahead
                    cache     = "_cache")     # optional; caching grib index

# Points of interest
library("sf")
locations <- data.frame(name = c("Innsbruck", "Brussels"),
                        lon  = c(11.39, 4.35),
                        lat  = c(47.27, 50.85))
(locations <- st_as_sf(locations, coords = c("lon", "lat"), crs = 4326))
```


## Getting data: NetCDF format {#gridded-download-netcdf}

The function `eupp_download_gridded()` allows to specify the `output_format =
"nc"` to store the data in the NetCDF file format. The procedure is, in large
parts, similar to retrieving the data in GRIB format (see [getting
started](gridded.html)). The function first downloads and stores the data in a
temporary GRIB file before calling
[`grib_to_netcdf`](https://confluence.ecmwf.int/display/ECC/grib_to_netcdf); a console
tool provided by ECMWFs [ecCodes](https://confluence.ecmwf.int/display/ECC/ecCodes+Home).

There seem to be ways to use `ecCodes` under Windows, see
[here](https://github.com/nawendt/gribr#windows-install-options) or
[here](https://www.ecmwf.int/en/newsletter/159/news/eccodes-and-magics-available-under-windows).

`grib_to_netcdf` is called with option `-k 3` resulting in a file with the
'netCDF-4 file format'.  Can be changed by adjusting the argument `netcdf_kind`
(default `3`, see `grib_to_netcdf` documentation).
Note that the conversion from GRIB to NetCDF adjusts the
variable names if needed (e.g., `2t` gets `t2m`).

```{r}
eupp_download_gridded(conf, output_file = "_test.nc", output_format = "nc", overwrite = TRUE)
```

These files can later be processed with the packages/software of your choice
(e.g., via [`ncdf4`][netcdf4], [`stars`][stars], CDS, ...).  Some examples how
to proceed from here ...

##### Reading data using `ncdf4`

```{r}
library("ncdf4")
nc <- nc_open("_test.nc", write = FALSE)
nc
```

##### Reading data using `stars`

```{r demo_read_stars, eval = FALSE}
library("stars")
x <- read_stars("_test.nc")
```

```{r demo_read_stars_hidden, include = FALSE}
library("stars")
x <- read_stars("_test.nc")
```

```{r gridded_demoplot_stars, fig = TRUE, fig.width = 10, fig.height = 5, out.width = "100%"}
print(x)
dim(x)
names(x)
st_get_dimension_values(x, "time") # date/time when valid

# Plotting 2m dry air temperature
plot(x["t2m"],
     main   = "t2m",
     col    = hcl.colors(21, "Blue-Red 2", rev = TRUE),
     breaks = seq(min(x$t2m, na.rm = TRUE), max(x$t2m, na.rm = TRUE), length = 22))

# Plotting convective precipitation (mm)
bk <- unique(quantile(x[["cp"]], p = seq(0, 1, by = 0.05)))
plot(x["cp"] * 1e3,
     main   = "cp",
     col    = hcl.colors(length(bk) - 1, "Purple-Yellow", rev = TRUE),
     breaks = bk * 1e3)
```

An alternative way to visualize the data using [`ggplot2`](https://cran.r-project.org/package=ggplot2),
[`colorspace`](https://cran.r-project.org/package=colorspace) and 
[`sf`](https://cran.r-project.org/package=sf) as well as 
[`rnaturalearth`](https://cran.r-project.org/package=rnaturalearth) for outlines:
```


```{r gridded_demoplot_ggplot, fig = TRUE, fig.width = 10, fig.height = 7, out.width = "60%", fig.align = "center"}
library("sf")
library("ggplot2")
library("colorspace")
library("rnaturalearth")

ne <- ne_countries(continent = "europe", returnclass = "sf")
st_crs(x) <- st_crs(ne)

# Plotting convective precipitation; member number 10.
# x["cp",,,number_idx,] is subsetting the stars object.
number_idx <- which(st_get_dimension_values(x, "number") == 10)
ggplot() + geom_stars(data = x["cp",,,number_idx,] * 1e3) +
           facet_wrap("time") +
           scale_fill_continuous_sequential("Purple-Yellow") +
           geom_sf(data = ne, fill = "transparent") +
           coord_sf(xlim = c(-6, 17), ylim = c(36, 67), expand = FALSE)
```

```{r demoplot_ggplot_lambert, fig = TRUE, fig.width = 10, fig.height = 7, out.width = "80%", fig.align = "center"}
# CRS ESRI:102014: Europe_Lambert_Conformal_Conic projection
# Hardcoded x/y-limits added; limit to southern Europe.
tmp <- st_transform(x, crs = st_crs("ESRI:102014"))
ggplot() + geom_stars(data = tmp["t2m",,,1,2]) +
           scale_fill_continuous_sequential("Red-Yellow") +
           geom_sf(data = ne, fill = "transparent") +
           coord_sf(xlim = c(-1.2e6, 6e5), ylim = c(7e5, 2e6), expand = TRUE)
```

## Getting data: `stars` {#gridded-get-stars}

As an alternative the package allows to directly retrieve the data as a [`stars`][stars]
object. Under the hood _eupp_ will download the gridded data set, converts it into 
a NetCDF file before calling `read_stars()`. Thus, [ecCodes][eccodes] will be required
as well and the variable names may differ from the original GRIB short name.

An example using the existing configuration `conf`:

```{r, results = "hidden"}
xst <- eupp_get_gridded(conf)
```

```{r gridded_get_stars, fig = TRUE, fig.width = 10, fig.height = 5, out.width = "100%", results = "hidden"}
class(xst)
names(xst)
print(xst)

plot(xst["t2m"], col = hcl.colors(11, "Green-Orange"))
```

To be precise, the object returned is of class ``r deparse(class(xst))`` which is
a pure `stars` object; the `eupp_stars` extension adds some additional functionality
for data handling (see `methods(class = "eupp_stars")`).



```{r gridded_cleaning_up, include = FALSE}
if (dir.exists("_cache")) unlink("_cache")
if (file.exists("_test.nc")) unlink("_test.nc")
if (file.exists("_test.grb")) unlink("_test.grb")
if (file.exists("_test.grb.rds")) unlink("_test.grb.rds")
```



[ecCodes]: https://confluence.ecmwf.int/display/ECC/ecCodes+Home
[stars]: https://cran.r-project.org/package=stars
[netcdf4]: https://cran.r-project.org/package=ncdf4
[rgdal]: https://cran.r-project.org/package=rgdal
