---
title: "Gridded Forecasts"
---

This manual shows how to download forecast data which includes both
[**high-resolution forecasts**](#hr) (the deterministic run) as well as
[**ensemble forecasts**](#ens) (there is a dedicated vignette for hindcasts/reforecasts).
It works the very same as for all other EUPP gridded
data sets, however, there are some more types available.

## High-resolution forecasts {#hr}

`type = "hr"` specifies downloading data from the high-resolution ECMWF
forecast run (deterministic run). Note that the data set is provided on the
same spatial resolution as all other gridded EUPP data sets ($0.25$ degrees
regular ll grid).

High-resolution forecasts are available on multiple levels:

* `"surface"`: surface fields (used as example in the following sections)
* `"pressure"`: pressure level data (see [here](#hr-pressure))
* `"efi"`": extreme forecast index data (see [here](#hr-efi))

### High-resolution surface data: Specify data

```{r, include = FALSE}
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("eupp"))
suppressPackageStartupMessages(library("stars"))
```

```{r, include = FALSE}
dir.create("_cache", showWarnings = FALSE) # Not mandatory
```

Step one is to set up before starting downloading data a configuration object
must be created using `eupp_config()` which contains the specification of the
data to be retrieved.

In this case high-resolution forecasts initialized on three different days,
only forecasts for `"2t"` (2m air temperature) and `"tcc"` (total cloud cover)
including forecasts 12h ahead up to 96h ahead from initialization in a
12-hourly interval.

```{r gridded_forecast_hr_config}
# Loading the package
library("eupp")

# Creating a vector of dates (character string in ISO representation works fine)
# specifying the date of model initialization (00 UTC).
dates <- c("2018-01-01", "2018-01-08", "2018-01-15")

# Create custom configuration
conf <- eupp_config(product   = "forecast",
                    level     = "surface",
                    type      = "hr",           # high-resolution forecast run
                    date      = dates,          # dates
                    parameter = c("2t", "tcc"), # single parameter
                    steps     = seq(12, 96, by = 12), # 12-hrly +12 to +96
                    cache     = "_cache")       # optional
```

The function returns an object of class `eupp_config` containing the
data of interest used for downloading the data in a second step.


### High-resolution surface data: Download GRIB

For all gridded EUPP data sets three options for downloading the data
exist. `eupp_download_gridded()` allows to download and store
the data set on disc.


```{r gridded_forecast_hr_download, eval = FALSE}
eupp_download_gridded(conf, "_hr_surface.grb", "grib", overwrite = TRUE)
```


### High-resolution surface data: Interpolation

```{r gridded_forecast_hr_processing, out.width = "100%", fig.width = 10, fig.height = 6}
library("sf")
locations <- data.frame(name = c("Innsbruck", "Brussels"),
                        lon  = c(11.39, 4.35),
                        lat  = c(47.27, 50.85))
(locations <- st_as_sf(locations, coords = c("lon", "lat"), crs = 4326))

# Perform bilinear interpolation
ip <- eupp_interpolate_grib("_hr_surface.grb", at = locations,
                            atname = "name", bilinear = TRUE)
head(ip, n = 4)
```



### High-resolution pressure level data

Only minor differences to surface data. All needed to be done is to 
set `level = "pressure"` and specify what parameter(s) to be downloaded.
Note: There is no dedicated option to specify which levels An example:

```{r gridded_forecast_hr_pressure, fig = TRUE, fig.width = 8, fig.height = 5, out.width = "80%"}
conf <- eupp_config("forecast", "pressure", "hr",
                    date = "2017-01-01",
                    steps = 12,
                    cache = "_cache")

eupp_download_gridded(conf, "_hr_pressure.grb", overwrite = TRUE)

ip <- eupp_interpolate_grib("_hr_pressure.grb", at = locations,
                            atname = "name", bilinear = TRUE)
head(ip, n = 4)
```


### High-resolution extreme forecast index

The extreme forecast index is a second-level product derived from the entire
atmosphere of all ensemble forecasts projected to a 2D plane. Thus, there are
no members or dedicated levels and only available aggregated over 24 hour time
periods available at 00 UTC (e.g., +0-24 hours ahead, +24-48 hours ahead, ...).

To get the data simply ask for `level = "efi"` (here not declaring `steps` to get all available forecast steps).

```{r gridded_forecast_efi, fig = TRUE, fig.width = 8, fig.height = 5, out.width = "80%"}
conf <- eupp_config("forecast", "efi",
                    date = "2017-01-01",
                    cache = "_cache")

eupp_download_gridded(conf, "_hr_efi.grb", overwrite = TRUE)

ip <- eupp_interpolate_grib("_hr_efi.grb", at = locations,
                            atname = "name", bilinear = TRUE)
head(ip, n = 4)
```

Note that not all parameters are available for all forecast steps (mainly
for the very long range horizonts).



## Ensemble forecasts {#ens}

`type = "ens"` specifies downloading data from the operational ensemble.
Works the same as for high-resolution forecasts (deterministic run) except
an optional argument `members` is available. The argument allows to specify
which ensemble forecast members/perturbations shall be processed.

* `0`: identifies the control run (treated the same as the perturbations)
* [`1`, ..., `50`]: individual perturbations
* If nothing is specified (`members = NULL`), the control run plus all
  available ensemble members will be downloaded/processed.


### Ensemble surface data

As always, `eupp_config()` has to be called to specify what should be processed
in the next step.

In this case ensemble forecasts initialized on three separate days (00 UTC)
for one single forecast parameter (`"cp"` = convective precipitation; `"surface"`
variable) valid at a forecast horizon of +120 hours.

In addition, `members = 0:3` defines to only download gridded data for the
control run (`0`) and the first three perturbations.

```{r gridded_forecast_ens_config}
# Loading the package
library("eupp")

# Create custom configuration
conf <- eupp_config(product   = "forecast",
                    level     = "surface",
                    type      = "ens",          # high-resolution forecast run
                    date      = "2017-01-02",   # dates
                    parameter = "cp",           # one parameter
                    steps     = 72:120,         # All available in this range
                    members   = 0:3,            # control run + members 1, 2, 3
                    cache     = "_cache")       # optional
```


Once the configuration object has been created we can download the data.

```{r gridded_forecast_ens_download, eval = FALSE}
eupp_download_gridded(conf, "_ens_surface.grb", "grib", overwrite = TRUE)
```

Last but not least interpolate the data using `eupp_interpolate_grib()`.
Note that, as we have ensemble forecasts, there are multiple forecast members
for the same date/time and parameter. This will be reflected by the adjusted
parameter names. `_[0-9]+$` at the end indicates the different members.

```{r}
ip <- eupp_interpolate_grib("_ens_surface.grb", at = locations,
                            atname = "name", bilinear = TRUE)
head(ip, n = 2)
```

The same is true when using the long format except .. its the long format.

```{r}
ip2 <- eupp_interpolate_grib("_ens_surface.grb", at = locations,
                             atname = "name", bilinear = TRUE,
                             wide = FALSE)
head(ip2, n = 3)
```

```{r ens_surface_lineplot, fig = TRUE, fig.width = 10, fig.height = 5, out.width = "100%"}
library("ggplot2")
ggplot(ip2) + geom_line(aes(x = valid, y = value, color = param)) +
    facet_wrap("name")
```

@TODO: Still cumulated!


### Ensemble pressure level data


```{r gridded_forecast_ens_pressure_config}
# Create custom configuration
conf <- eupp_config(product   = "forecast",
                    level     = "pressure",
                    type      = "ens",          # high-resolution forecast run
                    date      = "2017-01-02",   # dates
                    steps     = 72,             # Only forecast step 72
                    members   = 10:11,          # control run + members 10 and 11
                    cache     = "_cache")       # optional

k <- eupp_download_gridded(conf, "_ens_pressure.grb", overwrite = TRUE)

(ip <- eupp_interpolate_grib("_ens_pressure.grb", at = locations,
                             atname = "name", bilinear = TRUE))
```



