---
title: "Gridded data: Advanced"
---

```{r, include = FALSE}
suppressPackageStartupMessages(library("eupp"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("stars"))
dir.create("_cache", showWarnings = FALSE)
```

## Underlying concept in more details

The article "[_Getting started with gridded data_](gridded.html#underlying-concept)"
shows the basic use of the _eupp_ gridded data interface. For those interested, this
article shows some more insights how the package works under the hood.

Some functions might be useful for (i) debugging or (ii) adding additional functionality
around the _eupp_ package. In some more detail, the gridded dataset functionality
works as follows:

1. The user specifies the data set to be downloaded/retrieved using
   the `eupp_config()` function which returns an object of class `eupp_config`.
2. The user calls `eupp_download_gridded()` or `eupp_get_gridded()` to retrieve
   the data in different formats (the first allows for GRIB version 1 and different
   NetCDF file formats; the latter for `stars` objects). Below the surface _eupp_
   performs the following steps:
    a.  Defining the GRIB index files required to identify the necessary GRIB messages
    b.  Downloading and parsing the GRIB index files to identify files and byte ranges
    c.  Partially downloading the GRIB files (required messages via `curl`) and stores
        the requested messages in a new GRIB version 1 file.
    d.  If a NetCDF file has been requested: making required manipulations on the GRIB
        file and converting it to NetCDF, wherefore [ecCodes][ecCodes] needs to be installed.
    e.  If a [`stars`](https://cran.r-project.org/package=stars) object has been requested:
        read the NetCDF file. This goes trough the intermediate step of creating a NetCDF
        file; thus [ecCodes][ecCodes] is necessary.

When calling `eupp_download_gridded()` a file will be created on success (GRIB version 1 or
NetCDF), while `eupp_get_gridded()` returns a `stars` object in the active _R_ session.
Temporary files are deleted as soon as no longer needed (stored in `tempdir()`).

## Specify dataset to be downloaded

To demonstrate the intermediate steps listed above, a data set specification
(configuration) is required. For this purpose a small subset of gridded
surface ensemble forecast data is used.

* `cache = "_cache"`: Enables GRIB index caching which can be useful
  if the same GRIB indes file has to be accessed multiple times (as in this article).
* Imagine not knowing which parameters, forecast steps, or perturbation numbers (members)
  are available.

```{r gridded_advanced_config}
library("eupp")
(conf <- eupp_config(product   = "forecast",                    # forecasts
                     type      = "ens",                         # ensemble forecasts
                     level     = "surface",                     # surface fields
                     date      = c("2017-05-05", "2017-06-05"), # 'random' dates; ISO YYYY-mm-dd
                     cache     = "_cache"))                     # enable caching
```

Until now an _R_ object of class `eupp_config` has been created which is used
further down in the pipeline to process the request.


## Define required files

The next step is to define the URL(s) of the file(s) to be accessed to
process the request. This is done by the function `eupp_get_source_urls()`.

```{r}
# Required GRIB index files:
eupp_get_source_urls(conf, fileext = "index")
```

As shown above, four different files have to be accessed as we (i) are asking
for forecasts issued on two different dates (`date`) and have not explicitly defined
`members` wherefore we need both, control run forecasts (handled as `member = 0`) and
perturbed forecasts (`members` 1, 2, ...).

When `fileext` is not defined (`fileext = NULL`; default) one gets the URLs for the
corresponding GRIB files for direct access.


## Getting (full) inventory {#gridded-advanced-inventory}

In this scenario we imagined not having more information on what is available.
To get more insights we can use the configuration `conf` from above to get a complete
list of all messages in the GRIB index inventories listed above by calling
`eupp_get_inventory()`.

`eupp_get_inventory()` internally calls `eupp_get_source_urls(..., fileext = "index")`,
downloads the index files (line-wise JSON strings), parses them, and puts them into
an object of class `c("eupp_inventory", "data.frame")` (basic `data.frame`; no dedicated S3 methods so far).

```{r}
# Getting inventory (based on `conf` from above)
inv <- eupp_get_inventory(conf)
class(inv)
dim(inv)
head(inv)
```

As `cache` is enabled, the resulting `data.frame` is stored in _R_s RDS file format
into the `cache` folder; using an md5 checksum of the original URL to keep track of the
origin. When downloading another set of data stored in the same GRIB file (thus, same GRIB index file)
the cached file will be used which can significantly increase the performance.

The object returned contains information about the `path` of the grib file (not full URL) alongside
with a series of additional information which differ between different products. This inventory tells
us that the following parameters (`param`), steps (`step`), and ensemble members (`number`; perturbation number)
are available.

```{r}
unique(inv$param)
unique(inv$step)
unique(inv$number)
```

The full URL to the grib files can be constructed based on `inv$path` and `$BASEURL` from
`eupp:::eupp_get_url_config()` (can be redefined using system environment variable `EUPP_BASEURL`).
`eupp:::eupp_get_url_config()` not only returns the `BASEURL` but a series of template strings for the
different files on the bucket.



## Refine data set specification

A more practical use is to more explicitly define the data set configuration (as we now know what's needed).
Given `cache` was used above, the GRIB index file should be loaded from disc in a few secs.

```{r gridded_advanced_config2}
library("eupp")
(conf <- eupp_config(product   = "forecast",
                     type      = "ens",
                     level     = "surface",
                     date      = c("2017-05-05", "2017-06-05"),
                     parameter = c("tp", "sd"),                 # total precip + sunshine duration
                     steps     = seq(13, 15, by = 2L),          # +13 and +15 hour ahead forecast
                     members   = c(10, 14),                     # perturbation 10 and 14 (why not)
                     cache     = "_cache"))                     # use caching
```

Getting the required part of the inventory given the configuration above:

```{r}
(inv <- eupp_get_inventory(conf))
dim(inv)
```

The number of observations (rows) in `conf` matches our exception as asking for
(i) two different initialization dates, (ii) two parameters, (iii) two forecast
steps (lead times), and (iv) two different members ($2^4 = 16$).


## Downloading data

The data sets can be retrieved in three different formats which, however, are connected
(top down).

1. GRIB version 1 (minimal requirements; `curl`/`rcurl`)
2. NetCDF (requires [ecCodes][ecCodes] to be installed)
3. `stars` (requires the _stars_ package plus [ecCodes][ecCodes])

Given the inventory above the _eupp_ package first downloads segments of the original GRIB file
via curl byterange. The result is stored in one GRIB file. If this what has been requested by the
user, that's it (1). If the user requests a NetCDF file the GRIB file is stored temporarily and then
converted to NetCDF (the console tool `grib_set` is used to perform some ensemble-required manipulations;
then converted to NetCDF via `grib_to_netcdf`). When asking for `stars` objects we go trough the two steps
above before reading the data sets via `read_stars()` (_stars_ package). The conversion GRIB > NetCDF > stars
is required to do some naming manipulation.


**Download data as GRIB Version 1:**

* `eupp_download_gridded(conf, "my_new_file.grib", "grib")`
* See also: [Getting started: Getting data: GRIB format](gridded.html#gridded-download-grib)

**Download data and store as NetCDF:**

* `eupp_download_gridded(conf, "my_new_file.grib", "grib")`.
* _Note_: `netcdf_kind` can be used to control
  the `-k` flag when calling `grib_to_netcdf` (defaults to `3`);
  see `eupp_download_gridded()` and `grib_to_netcdf` manual.
* See also: [Getting started: Getting data: NetCDF format](gridded.html#gridded-download-netcdf)

**Getting data as [`stars`][stars] object:**

* `x <- eupp_get_gridded(conf)`
* See also: [Getting started: `stars`](gridded.html#gridded-get-stars)


























[ecCodes]: https://confluence.ecmwf.int/display/ECC/ecCodes+Home
