---
title: "Interpolating data"
---

```{r, include = FALSE}
# Just here to suppress startup message
for (x in c("eupp", "stars", "sf")) suppressPackageStartupMessages(x)
```


Loading the packages required to perform this action:

```{r}
library("eupp")  # for downloading/processing data
library("sf")    # for point locations
library("stars") # for interpolating data (st_extract)
```

### Setting up an `[sf][sf]` object

Step 1, set up an `[sf][sf]` `data.frame` with point locations (`sf_point`)
which will be used for interpolation. The `name` is not per se required,
but used to further process the data.

In this demo we are reading a CSV file from an object. This can be done
in many different ways; check out the `[sf][sf]` package to find out more.

```{r}
locations <- "name, lon, lat
Innsbruck,11.39333,47.26833
Vienna,16.37255,48.20833
Bonn,7.09981,50.73399
Reading,-0.97305,51.45416"

(locations <- st_as_sf(read.csv(text = locations), coords = c("lon", "lat")))
```

### Downloading EUPP data (surface analysis)

For demonstration we will download surface analysis data from the
EUPP data set for one specific date. Parameters `"2t"` (2m above
ground dry air temperature) and `"cp"` (convectie precipitation) for
every hour of the day (`step = NULL`).

`eupp_get_gridded()` will return an object of class `c("eupp_stars", "stars")`
(see `?eupp_stars`) which provides some convenience methods to process the
data while, internally, being a barebone `stars` object.


```{r}
config <- eupp_config("analysis", "surface",
                      date = "2017-03-01",
                      parameter = "2t", cache = "_cache")
ana <- eupp_get_gridded(config)
```

### Interpolating data

Given we have our `c("eupp_stars", "stars")` object and the
`c("sf", "data.frame")` object (`locations`) we can perform
interpolation using the `[sf_extract()][sf]` function.

The function returns a `stars` object which can be coerced
into a `data.frame`.

For plotting, we will append the `name` of the location again.

```{r}
res <- st_extract(ana, locations, bilinear = TRUE) |> as.data.frame()
# Appending location name
res$name <- locations$name[match(res$geometry, locations$geometry)]
head(res)
```

### Plotting

```{r, out.width = "100%", fig.width = 12, fig.height = 5}
library("ggplot2")
ggplot(res) + geom_line(aes(x = time, y = t2m - 273.15, group = name, colour = name))
```


[sf]: https://cran.r-project.org/package=sf
